Automating Political Bias Prediction

### 2016-03-31 manuscript/political_bias_prediction.tex
yet more text

Ideally one would choose for each topic a sample of reports from the entire political spectrum in order to form an unbiased opinion. But ordering and assigning media content with respect to the political spectrum at scale requires automated prediction of political bias. The aim of this study is to provide some empirical evidence indicating that political bias prediction is possible with well above chance accuracy.

@@

\section{Classification Model and Training Procedure}

@@

Next to the four class problem of categorizing all four parties of the german parliament we also considered the binary classification task of only distinguishing texts between wether they were read by members of the government or members of the opposition. Results of this simpler two-class problem were obtained by exactly the same methods as in the four class case.

\subsection{Optimisation of Model Parameters}\label{sec:crossvalidation}
The model pipeline contained a number of {\em hyperparameters} that we did not want to tune by hand. These parameters are called hyperparameters becase we cannot optimize these like we optimize the parameters of our classifier, by writing down an objective function and optimizing this linear function by doing gradient descent. Thus we resort to the standard 'trial-and-error' approach in machine learning and perform cross-validation on the hyperparameters. In order to not overfit to our data and obtain good generalization performance to other texts we used two approaches. The first one is the usual procedure of 'nested-crossvalidation', which performs two stages of cross-validation, an {\em outer} cross-validation loop to evaluate the model correctly on held-out data, and an {\em inner} cross-validation loop within each training data fold of the outer cross-validation. In practice this means that in each step of the outer loop we set aside a part of our data set and then try out all parameters using the inner cross-validation on the not-held-out data, and afterwards make a prediction on the held-out data of the outer cross-validation. For each set of preprocessing (hyper-) parameters and regularisation parameter of the classifier, we performed the entire data extraction, vectorization and classifier training workflow. Next to this rather standard approach, we also tested the classifiers on the text of the party programs (see \autoref{sec:data}) which were never included in any of the training procedures.

@@

\section{Results}

@@

\section{Speech sentiment correlates with political power}\label{sec:sentiment_result}
We analysed the sentiment of each speech of a political party and plot the results in \autoref{fig:party_sentiments}. While the methodology used for estimating the sentiment (see \autoref{sec:sentiment_analysis_methods}), it can yield some insight into the emotional content of a speech. High values indicate more pronounced usage of positive words, whereas negative values indicate more pronounced usage of words associated with negative emotional content. We can quantify political power as the number of seats a party has in the parliament. Our analyses show a strong trend supporting the hypothesis that the more political power a party has, the more positive sentiment their speech has, and vice versa, the less seats a party has, the more negative content their speeches contain. This pattern is evident from the barplot in \autoref{fig:party_sentiments} and supported by a high correlation of 0.86 between number of parliament seats and sentiment index.

@@

\section{Correlations between words and parties}\label{sec:word_party_correlations}

@@

>>>>

Ideally one would choose for each topic a sample of reports from the entire political spectrum in order to form an unbiased opinion. But ordering and assigning media content with respect to the political spectrum at scale requires automated prediction of political bias. The aim of this study is to provide some empirical evidence indicating that political bias prediction is possible with well above chance accuracy. \\

Automated predictions of political bias can be problematic for ethical reasons. One trusts political experts, as they will take responsibility for what they say about a text, and they can explain their decisions. This is a key difference to many statistical learning approaches. Not only is the responsibility question not entirely clear, it can also be difficult to interpret some of the decisions. In order to validate the predictions of the models we devise two strategies that allow for better interpretations of the models. First we are looking at univariate measures of how discriminative the features are that the model uses. This somewhat allows to relate the political texts to the kind of information that political experts use, too. Second we use sentiment analysis to investigate whether this aspect of language also has discriminatory power. \\

In the following \autoref{sec:data} gives an overview of data acquisition and preprocessing, \autoref{sec:model} presents the model, in \autoref{sec:results} the results are discussed and \autoref{sec:conclusion} concludes with some interpretations of the results and future research directions.

@@

\section{Classification Model and Training Procedure}\label{sec:model}

@@

We considered three different classification problems: 1) classification of party affiliation (5 class problem for the 17th legislation period, 4 class problem for the 18th legislation period), 2) classification of government membership (binary problem) and 3) classification of political views as expressed by the political codes of the manifestoproject (56 class problem). For each of first two problems, party affiliation and government membership prediction, we trained classifiers on the parliament speeches. For the third problem we trained only on the manifesto data as only for those the political view labels were available.

\subsection{Optimisation of Model Parameters}\label{sec:crossvalidation}
The model pipeline contained a number of {\em hyperparameters} that we did not want to tune by hand. 
We first split the training data into a training data set and a test data set for optimizing the hyperparameters, the test data for hyperparameter optimization was 10\% of the entire training data. Hyperparameters were optimized using grid search: For each setting of hyperparameters the entire pipeline was trained and evaluated on the test set. For the best setting of all hyperparameters, we trained the pipeline again on all training data and evaluated it on the evaluation set. Evaluation data was never used for training the models. For party affiliation prediction and government membership prediction the training and test set were 90\% and 10\%, respectively, of all data in a legislative period. Evaluation data were the texts from party manifestos. For the political view prediction setting we first split all labeled manifesto sentences in both legislative periods into a training/test and evaluation set of 90\% (train/test) and 10\% (evaluation), then we split the train/test part again with a 90\%/10\% ratio for training/testing of the hyperparameters.

@@

\section{Results}\label{sec:results}

@@

\subsection{Speech sentiment correlates with political power}\label{sec:sentiment_result}
We analysed the sentiment of each speech of a political party and plot the results in \autoref{fig:party_sentiments}. While the methodology used for estimating the sentiment (see \autoref{sec:sentiment_analysis_methods}), it can yield some insight into the emotional content of a speech. High values indicate more pronounced usage of positive words, whereas negative values indicate more pronounced usage of words associated with negative emotional content. We can quantify political power as the number of seats a party has in the parliament. Our analyses show a strong trend supporting the hypothesis that the more political power a party has, the more positive sentiment their speech has, and vice versa, the less seats a party has, the more negative content their speeches contain. This pattern is evident from the barplot in \autoref{fig:party_sentiments} and supported by a high correlation of 0.86 between number of parliament seats and sentiment index.

@@

\begin{table}[t]
\begin{center}
\begin{tabular}{lrr}
   Sentiment vs. &          Gov. Member    &  Seats\\
\hline\hline
17th period    &  0.84 & 0.70\\
18th period    &  0.98 & 0.89\\
%
\end{tabular}
\end{center}
\caption{
\label{tab:results_avg_political_view}
Correlation coefficient between average sentiment of political speeches of a party in the german bundestag with two measures of their political power, a) their membership in the government and b) with the number of seats that party occupies in the parliament. Shown are correlations for the 17th legislative period and the (ongoing) 18th legislative period.
}
\end{table}

\subsection{Correlations between words and parties}\label{sec:word_party_correlations}

@@

\section{Conclusions and Outlook}\label{sec:conclusion}
